TacticToe: Learning to Prove with Tactics

### 2018-03-14 tactictoe_jv.tex
spellcheck

\section{Introduction}
Many of the state-of-the-art interactive theorem provers (ITPs) such as
  \holfour~\cite{hol4}, \hollight~\cite{Harrison09hollight},
  \isabelle~\cite{isabelle}
  and \coq~\cite{coq-book} provide high-level parameterizable tactics for constructing proofs.
  Tactics analyze the current proof state (goal and
  assumptions) and apply non-trivial proof transformations.
  Formalized proofs take advantage of different levels of automation:
  specialized rules, theory-based strategies and general purpose startegies.
  Thanks to progress in proof
  automation, developers can delegate more and more complicated proof
  obligations to general purpose strategies. Those are implemented by automated
  theorem provers (ATPs) such as \eprover~\cite{eprover}. Communication
  between
  an ITP and ATPs is made possible by a ``hammer''
  system~\cite{hammers4qed,tgck-cpp15}. It acts as an interface by performing
  premise selection, translation and proof reconstruction.
  Yet, ATPs are far from foolproof and more precise user-guidance, achieved
  by applying a particular sequence of specialized rules, is almost always
  necessary to develop a mathematical theory.

@@

%A similar trade-off between exploration and exploitation exists in algorithm
%like Monte Carlo. However, Monte Carlo tree search metrics are based on many
%simulations and do not reflect our exploration strategy that only
\paragraph{Node selection}
Through node extension steps the search tree grows and the number of paths to
explore increases. To decide which node to extend next, an evolving value for
each node is given in Definition~\ref{def:value}. The algorithm that performs
node selection starts from the root of the search tree. From the current node,
the function \texttt{CompareChildren} choses among the children of its open
goal the one with the highest value.

@@

In Table~\ref{tab:tuning}, the success rate of the different \tactictoe
strategy is presented.
The first four techniques are tested relative to the same baseline indicated in
the table by the tag ``default''.  This default strategy relies on a tactic
time out of 0.05s, an orthogonalization radius of 20 and a policy coefficent of
0.5.
The two last techniques are tested relative
to a baseline consisting of the best set of parameters discovered so far and
are marked with the improvement over the ``default'' strategy.
In each subtable, each experiment differs from the current baseline by exactly
one parameter.

Thanks to the larger search time available in this experiment, the timeout for
each tactic can be increased from the 0.02 seconds used in \tactictoe's initial
experiments~\cite{tgckju-lpar17} to 0.05
seconds. Removing tactics with similar effect as performed by the
orthogonalization process is only beneficial when running \tactictoe for short
period of time. It seems that the alloted time allows a strategy
running without orthogonalization to catch up.
It is best to predict a list of 16 theorems to instantiate arguments of type
theorem list in tactics. At any rate, argument prediction for abstracted
tactics is the technique that have the highest impact on the success of
\tactictoe. Integration of ATPs being a specialization of this technique
contributes significantly as well. Experiments involving \eprover, run the ATP
using a separate proces asynchronously with a timeout of 5 seconds and 128 premises.
The fact that \metis outperforms \eprover in this setting is due to the fact
that \metis is run with a very short timeout, allowing to close different part
of search tree quickly. Because \metis is weaker that \tactictoe as an ATP, it
is given less predictions. But even if an essential lemma for the proof of a
goal is not predicted, a modification on the goal performed by a tactic can
change its prediction to include the necessary lemma.
This effect minimizes the drawback of relying on a small number of predictions.

@@

The extracted proof script of $\mathfrak{T}$ is $P_{goal}(g_{root})$.
We minimally improve it by subtituting \texttt{THENL} by \texttt{THEN} when the
list of goals is a singleton and removing \texttt{THENL []} during the
extraction phase.
Further post-processing such as
eliminating unnecessary tactics and theorems has been developed and
improves the user experience greatly~\cite{DBLP:conf/sefm/Adams15}.

@@

We start with Example~\ref{ex:cs1}, which proves that greatest common divisors are unique.
The human formalizer recognized that it follows from two theorems.
The relevance filtering of \tactictoe is not as powerful as that used in hammers
and is therefore not able to find the \texttt{DIVIDES\_ANTISYM} property. The proof
proceeds instead by rewriting the
goal with the definitions then spliting the goal into multiple cases until the goal obligations are similar
to \texttt{DIVIDES\_ANTISYM}. As expected, \tactictoe's proof
also takes much longer to check.

>>>>

\section{Introduction}
Many of the state-of-the-art interactive theorem provers (ITPs) such as
  \holfour~\cite{hol4}, \hollight~\cite{Harrison09hollight},
  \isabelle~\cite{isabelle}
  and \coq~\cite{coq-book} provide high-level parameterizable tactics for constructing proofs.
  Tactics analyze the current proof state (goal and
  assumptions) and apply non-trivial proof transformations.
  Formalized proofs take advantage of different levels of automation:
  specialized rules, theory-based strategies and general purpose strategies.
  Thanks to progress in proof
  automation, developers can delegate more and more complicated proof
  obligations to general purpose strategies. Those are implemented by automated
  theorem provers (ATPs) such as \eprover~\cite{eprover}. Communication
  between
  an ITP and ATPs is made possible by a ``hammer''
  system~\cite{hammers4qed,tgck-cpp15}. It acts as an interface by performing
  premise selection, translation and proof reconstruction.
  Yet, ATPs are far from foolproof and more precise user-guidance, achieved
  by applying a particular sequence of specialized rules, is almost always
  necessary to develop a mathematical theory.

@@

%A similar trade-off between exploration and exploitation exists in algorithm
%like Monte Carlo. However, Monte Carlo tree search metrics are based on many
%simulations and do not reflect our exploration strategy that only
\paragraph{Node selection}
Through node extension steps the search tree grows and the number of paths to
explore increases. To decide which node to extend next, an evolving value for
each node is given in Definition~\ref{def:value}. The algorithm that performs
node selection starts from the root of the search tree. From the current node,
the function \texttt{CompareChildren} chooses among the children of its open
goal the one with the highest value.

@@

In Table~\ref{tab:tuning}, the success rate of the different \tactictoe
strategy is presented.
The first four techniques are tested relative to the same baseline indicated in
the table by the tag ``default''.  This default strategy relies on a tactic
time out of 0.05s, an orthogonalization radius of 20 and a policy coefficient of
0.5.
The two last techniques are tested relative
to a baseline consisting of the best set of parameters discovered so far and
are marked with the improvement over the ``default'' strategy.
In each subtable, each experiment differs from the current baseline by exactly
one parameter.

Thanks to the larger search time available in this experiment, the timeout for
each tactic can be increased from the 0.02 seconds used in \tactictoe's initial
experiments~\cite{tgckju-lpar17} to 0.05
seconds. Removing tactics with similar effect as performed by the
orthogonalization process is only beneficial when running \tactictoe for short
period of time. It seems that the allotted time allows a strategy
running without orthogonalization to catch up.
It is best to predict a list of 16 theorems to instantiate arguments of type
theorem list in tactics. At any rate, argument prediction for abstracted
tactics is the technique that have the highest impact on the success of
\tactictoe. Integration of ATPs being a specialization of this technique
contributes significantly as well. Experiments involving \eprover, run the ATP
using a separate process asynchronously with a timeout of 5 seconds and 128 premises.
The fact that \metis outperforms \eprover in this setting is due to the fact
that \metis is run with a very short timeout, allowing to close different part
of search tree quickly. Because \metis is weaker that \tactictoe as an ATP, it
is given less predictions. But even if an essential lemma for the proof of a
goal is not predicted, a modification on the goal performed by a tactic can
change its prediction to include the necessary lemma.
This effect minimizes the drawback of relying on a small number of predictions.

@@

The extracted proof script of $\mathfrak{T}$ is $P_{goal}(g_{root})$.
We minimally improve it by substituting \texttt{THENL} by \texttt{THEN} when the
list of goals is a singleton and removing \texttt{THENL []} during the
extraction phase.
Further post-processing such as
eliminating unnecessary tactics and theorems has been developed and
improves the user experience greatly~\cite{DBLP:conf/sefm/Adams15}.

@@

We start with Example~\ref{ex:cs1}, which proves that greatest common divisors are unique.
The human formalizer recognized that it follows from two theorems.
The relevance filtering of \tactictoe is not as powerful as that used in hammers
and is therefore not able to find the \texttt{DIVIDES\_ANTISYM} property. The proof
proceeds instead by rewriting the
goal with the definitions then splitting the goal into multiple cases until the goal obligations are similar
to \texttt{DIVIDES\_ANTISYM}. As expected, \tactictoe's proof
also takes much longer to check.
